
# Data 



## Sources

We collect data from WHO Covid-19 Dashboard and CDC Data for Covid-19. We download the csv files from the websites.


### WHO Covid-19 Dashboard
There is a csv file for the cumulative cases and deaths data of each country. The country names, iso number, date reported are also included in the csv file. There are 239 records of countries and regions, including Global category and others category.

There is a csv file for global vaccination data. We can load the number of persons vaccinated with different numbers of vaccines and vaccination rates of different countries from the file. The country names, WHO region, iso number, date reported are also included in the csv file. There are 229 records of countries and regions.

### CDC Data for Covid-19



## Cleaning / transformation


### Covid-19 Cases and Deaths

#### Distribution by Space
The cumulative cases and deaths data from WHO can be loaded from the csv files directly. Then we select infection rates, fatality rates, country names and country iso numbers from the data frame. For the vaccine data, we load the csv file and select country names, country numbers and persons with 1+ doses of 100 population from the data frame. Then we store the data frames in Rdata files for future use. 


```{r}
library("tidyverse")
library("dplyr")
casedata <- read.csv("WHO-COVID-19-global-table-data.csv")
code<-read.csv("WHO-COVID-19-global-data.csv")
ccode=code%>%select(c("Country_code","Country"))
countrycode=ccode[!duplicated(ccode$Country),]
case_code=merge(casedata,countrycode,by.x="Name",by.y="Country")
df_case = subset(case_code, select = c(Country_code,Cases...cumulative.total,Cases...cumulative.total.per.100000.population,Deaths...cumulative.total,Deaths...cumulative.total.per.100000.population) )
#df_case:case and death with ISO Alpha-2 country code

save(df_case, file = "df_case.Rdata")
vaccinedata<- read.csv("vaccination-data.csv")
df_v=subset(vaccinedata,select=c(ISO3,PERSONS_VACCINATED_1PLUS_DOSE_PER100,PERSONS_FULLY_VACCINATED_PER100))
save(df_v, file = "df_v.Rdata")
```

### Distribution by Time

To analyze the global spread of Covid-19, it's helpful to observe the reported cases and deaths as a three-year time series from early 2020 to late 2020. The data is collected from the report of different countries separately, so we first sum them up to obtain the global reported data by day. We notice that there is a delay in data collection on weekends in the data set, which may lead to an undesired cyclic pattern. To deal with that, we also build another data frame by grouping the reported date by week. We believe this weekly reported data frame may bring clearer insights of the data.

We also extract the US data from the global data set, and compute the reported cases and deaths by day and week. It will provide a deeper insight in our following analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
who_daily <- read.csv("https://raw.githubusercontent.com/5702-Final-Project-7/Covid-19-Data-Analysis/main/WHO-COVID-19-global-data.csv")
# summarize the daily reported data of all the countries
world_daily <- who_daily %>%
  group_by(Date_reported) %>%
  mutate(world_new_cases = sum(New_cases)) %>%
  mutate(world_new_deaths = sum(New_deaths)) %>%
  select(c(Date_reported, world_new_cases, world_new_deaths)) %>%
  distinct()  %>%
  mutate(Date_reported = as.Date(Date_reported))


# summarize the weekly reported data to remove cyclic patterns
world_weekly <- world_daily %>%
  mutate(week = as.numeric(floor((Date_reported-as.Date('2020-01-03'))/7))) %>%
  group_by(week) %>%
  mutate(weekly_new_cases = sum(world_new_cases)) %>%
  mutate(weekly_new_deaths = sum(world_new_deaths)) %>%
  mutate(date_reported = as.Date(as.Date('2020-01-03')+week*7)) %>%
  select(date_reported,weekly_new_cases,weekly_new_deaths,week) %>%
  ungroup() %>%
  distinct()

# us daily reported data
us_daily <- who_daily %>%
  filter(Country == 'United States of America') %>%
  distinct() %>%
  select(Date_reported, New_cases, New_deaths) %>%
  mutate(Date_reported = as.Date(Date_reported)) %>%
  filter(New_deaths >= 0)

# summarize us weekly reported data
us_weekly <- us_daily %>%
  mutate(week = as.numeric(floor((Date_reported-as.Date('2020-01-03'))/7))) %>%
  group_by(week) %>%
  mutate(weekly_new_cases = sum(New_cases)) %>%
  mutate(weekly_new_deaths = sum(New_deaths)) %>%
  mutate(Date_reported = as.Date(as.Date('2020-01-03')+week*7)) %>%
  select(Date_reported,weekly_new_cases,weekly_new_deaths,week) %>%
  ungroup() %>%
  distinct()
```


### US Vaccination Data

#### US Vaccination Compared to the Global Data

To analyze the US vaccination rate, we tried to compare it to those of other countries in the world. We used the WHO global vaccination data, which provides detailed vaccination data divided by countries. For each country, it gives the total vaccinations, the vaccinated rate, fully vaccinated rate as well as the main type of vaccines the country uses. In our analysis, we only care about the vaccinated rate and fully vaccinated rate of different countries. Another issue of the data is missing values and outliers. There's no fully vaccinated rate for Guernsey, and the fully vaccinated rate of Eritrea, Brunei Darussalam, Gibraltar, Palau and Tokelau are more than 100%. So we drop these six rows. 


```{r}
world_vac_df <- read.csv('https://raw.githubusercontent.com/5702-Final-Project-7/Covid-19-Data-Analysis/main/vaccination-data.csv')
world_vac <- world_vac_df %>%
  select(COUNTRY,PERSONS_FULLY_VACCINATED_PER100) %>%
  rename(country=COUNTRY, fully_vaccinated_rate = PERSONS_FULLY_VACCINATED_PER100)

# check and remove missing values and outliers
na_index <- c(which(is.na(world_vac$fully_vaccinated_rate)),which(world_vac$fully_vaccinated_rate<=0))
outlier_index <- which(world_vac$fully_vaccinated_rate>100)
world_vac[c(na_index,outlier_index),]
world_vac <- world_vac[-c(na_index,outlier_index),]

```

#### Vaccination State in US Populations

We further made use of the global vaccination data set by extracting the US data, and calculating the proportion of vaccination, full vaccination, boosting in the US population. The data frame consists of 5 columns: 'country', 'not_vaccinated', 'vaccinated', 'fully_vaccinated', and 'boosted'. We transformed it to tidy form which only contains 3 columns for further graphing.


```{r}
us_vac <- world_vac_df %>%
  filter(COUNTRY=='United States of America') %>%
  select(PERSONS_VACCINATED_1PLUS_DOSE_PER100,
         PERSONS_FULLY_VACCINATED_PER100,
         PERSONS_BOOSTER_ADD_DOSE_PER100) %>%
  rename(vaccinated=PERSONS_VACCINATED_1PLUS_DOSE_PER100,
         fully_vaccinated=PERSONS_FULLY_VACCINATED_PER100,
         boosted=PERSONS_BOOSTER_ADD_DOSE_PER100) %>%
  mutate(not_vaccinated=100-vaccinated) %>%
  mutate(vaccinated=vaccinated-fully_vaccinated) %>%
  mutate(fully_vaccinated=fully_vaccinated-boosted) %>%
  mutate(country='United States of America') %>%
  pivot_longer(!country)
```


#### US Vaccination by State

Brief explanation here.

```{r echo=TRUE, message=FALSE, warning=FALSE}
vacraw <- read_csv("https://raw.githubusercontent.com/5702-Final-Project-7/Covid-19-Data-Analysis/main/covid19_vaccinations_us.csv")
vac_us <- vacraw %>%
  select(`Jurisdiction (State/Territory) or Federal Entity`, 
         `Percent of total pop with a completed primary series`) %>%
  rename(state = `Jurisdiction (State/Territory) or Federal Entity`,
         fully_vaccinated_rate = `Percent of total pop with a completed primary series`)

vac_us[vac_us == "N/A"] <- NA
vaccinated_us <- na.omit(vac_us)
vaccinated_us[, c(2)] <- sapply(vaccinated_us[, c(2)], as.numeric)
```

#### Vaccination and Reported Cases

Brief explanation here.

```{r echo=TRUE, message=FALSE, warning=FALSE}
dosesraw <- read_csv("https://raw.githubusercontent.com/govex/COVID-19/master/data_tables/vaccine_data/global_data/time_series_covid19_vaccine_doses_admin_global.csv")

doses_1 <- dosesraw %>%
  select(-(UID:`Admin2`)) %>%
  select(-(Combined_Key)) %>%
  select(-(Population)) %>%
  rename(Long = `Long_`)

doses <- doses_1 %>%
  pivot_longer(-c(`Province_State`, `Country_Region`, Lat, Long),
               names_to = "date",
               values_to = "doses_n") %>%
  select(-c(Lat, Long)) %>%
  rename(province_state = `Province_State`,
         country_region = `Country_Region`) %>%
  mutate(date = ymd(date)) %>%
  group_by(country_region, date) %>%
  summarise(doses_n = sum(doses_n)) %>%
  ungroup()

new_dose <- doses %>%
  arrange(date) %>%
  group_by(country_region) %>%
  mutate(new_doses_n = doses_n - lag(doses_n, default = 0)) %>%
  ungroup() %>%
  as.data.frame(sapply(new_dose, function(x) gsub("\"", "", x)))

new_dose[is.na(new_dose)] <- 0

weekly_dose <- new_dose %>%
  mutate(week = as.numeric(floor((date-as.Date('2020-12-29'))/7))) %>%
  group_by(week) %>%
  mutate(weekly_new_doses = sum(new_doses_n)) %>%
  mutate(date = as.Date(as.Date('2020-12-29') + week * 7)) %>%
  ungroup() %>%
  distinct()
```


```{r}
caseraw <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

case <- caseraw %>%
  pivot_longer(-c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "confirmed_n") %>%
  select(-c(Lat, Long)) %>%
  rename(province_state = `Province/State`,
         country_region = `Country/Region`) %>%
  mutate(date = mdy(date)) %>%
  group_by(country_region, date) %>%
  summarise(confirmed_n = sum(confirmed_n)) %>%
  ungroup()

new_case <- case %>%
  arrange(date) %>%
  group_by(country_region) %>%
  mutate(new_cases_n = confirmed_n - lag(confirmed_n, default = 0)) %>%
  ungroup()

weekly_case <- new_case %>%
  mutate(week = as.numeric(floor((date-as.Date('2020-01-22'))/7))) %>%
  group_by(week) %>%
  mutate(weekly_new_cases = sum(new_cases_n)) %>%
  mutate(date = as.Date(as.Date('2020-01-22') + week * 7)) %>%
  ungroup() %>%
  distinct()

weekly_merge <- merge(weekly_case, weekly_dose, by=c("country_region","date"), all=TRUE)
```


## Missing value analysis

### WHO Covid-19 dashboard

There are no NAN values in the cumulative cases and deaths data. But there are 4 countries/regions with 0 cases and deaths.They are Democratic People's Republic of Korea,Saint Helena,Tokelau and Turkmenistan. It implies that they didn't report to the WHO and their data are missing. There is a category "other" in the table and it may contain data of countries not included in the 237 countries/ regions.

There are 8 countries/regions which have cumulative cases and deaths data but don't have vaccination data. By comparing the two table we can find the countries are Bonaire, Sint Eustatius and Saba,Democratic Republic of the Congo,Holy See,Mayotte,Réunion,Saint Barthélemy,Saint Martin,and Saint Pierre and Miquelon. They are reltevely small regions which may lack relevant data.



```{r}
save.image(file='~/projectimage.Rdata')
```







