```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(scales)
library(lubridate)
library(glue)
library(ggridges)
library(colorspace)
library(ggpubr)
library("sf")
library("tidyverse")
library("leaflet")  
library("rnaturalearth")
library("rnaturalearthdata")
library("riskyr")
library("scales")
library("htmltools")
library(grid)
load(file='~/projectimage.RData')
```


# Results

## Global spatial data for cases & deaths
Here we present the global map of cumulative cases per 100000 population, deaths per 100000 population, and persons vaccinated 1+ doses per 100 population. The map is interactive. Put the mouse over a particular area, and the detailed data will appear. The maps can be zoomed in by clicking the buttons on the left and be dragged by holding the left button of the mouse. The data is updated on 2022/12/05.

We can see from the graphs that European countries show high infection rates, followed by the Americas. The reported infection rate in Asia and Africa is relatively late.  The bordering countries show similar infection rates, while the countries with strict covid-19 policies show lower rates compared to surrounding countries, for instance, China.  It has been noted on the WHO website that the case and death data for the Region of Africa are incomplete. Therefore the actual infection rates in Africa may be higher than reported.

Countries with high infection rates are more likely to report more deaths. For the vaccination data, we can see that countries in Europe and in the Americas tend to show higher vaccination rates, which are the richer countries with more medical resources.

We find from the data that the death rates are relatively high in the US compared to countries with similar or even higher infection rates in Europe. For example, the infection rate in the US is 29491/100000, and the death rate is 324/100000, while the figures for France are 566638/100000 and 238/100000. We first infer that it may be caused by the difference in calculating death cases. In the US, if COVID-19 appears among the causes and contributors, CDC guidance counts that as a COVID-19-related death. Some countries in Europe record solely deaths of those who have been tested and that occurred in the hospital. But for countries like France and German, their case numbers include those who died "of" COVID-19 and those who died "with" the disease, both in hospitals and care houses. The United Kingdom has been reporting all deaths after a positive test within 60 days as a coronavirus death. So the different counting methods in Europe may not be the main cause of the lower rates of deaths here. The reasons behind these differences may be complicated. From the data we have now, the US shows similar vaccination rates, and the differences may be caused by medical resources, various types of social distance measures, and time affected by Covid-19.

```{r}
load(file = "df_case.Rdata")
load(file = "df_v.Rdata")


world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)

worldmm<-merge(world,df_case,by.x="iso_a2",by.y="Country_code")
worldm<-merge(worldmm,df_v,by.x="iso_a3",by.y="ISO3")

palc <- colorBin("YlOrRd", domain = worldm$Cases...cumulative.total.per.100000.population)
labelsc <- sprintf(
  "<strong>%s</strong><br/> Cumulative cases per 100000: %s ",
  worldm$name, comma(worldm$Cases...cumulative.total.per.100000.population)) %>% 
  lapply(HTML)

pald <- colorBin("YlGn", domain = worldm$Deaths...cumulative.total.per.100000.population)
labelsd <- sprintf(
  "<strong>%s</strong><br/> Cumulative Deaths per 100000: %s ",
  worldm$name, comma(worldm$Deaths...cumulative.total.per.100000.population)) %>% 
  lapply(HTML)

palv <- colorBin("PuBu", domain = worldm$PERSONS_VACCINATED_1PLUS_DOSE_PER100,
                 bins=c(0,20,40,60,80,100))
labelsv <- sprintf(
  "<strong>%s</strong><br/> Persons vaccinated more than one doses per 100: %s ",
  worldm$name, comma(worldm$PERSONS_VACCINATED_1PLUS_DOSE_PER100)) %>% 
  lapply(HTML)




case_map<-leaflet(worldm) %>%
  addProviderTiles("Stamen.TonerLite")%>%
  addPolygons(
    # fill
    fillColor   = ~ palc(Cases...cumulative.total.per.100000.population),
    fillOpacity = 0.7,
    # line
    dashArray   = "3",
    weight      = 2,
    color       = "white",
    opacity     = 1,
    # interaction
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsc,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(
    pal = palc, values = ~Cases...cumulative.total.per.100000.population, opacity = 0.7, title = HTML("Cumulative cases per 100000 population"),
    position = "bottomright")
setView(case_map,lng=-30, lat=-20, 1)

```
```{r}
death_map<-leaflet(worldm) %>%
  addProviderTiles("Stamen.TonerLite")%>%
  addPolygons(
    # fill
    fillColor   = ~ pald(Deaths...cumulative.total.per.100000.population),
    fillOpacity = 0.7,
    # line
    dashArray   = "3",
    weight      = 2,
    color       = "white",
    opacity     = 1,
    # interaction
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsd,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(
    pal = pald, values = ~Deaths...cumulative.total.per.100000.population, opacity = 0.7, title = HTML("Cumulative deaths per 100000 population"),
    position = "bottomright")
setView(death_map,lng=-30, lat=-20, 1)


```
```{r}
v_map<-leaflet(worldm) %>%
  addProviderTiles("Stamen.TonerLite")%>%
  addPolygons(
    # fill
    fillColor   = ~ palv(PERSONS_VACCINATED_1PLUS_DOSE_PER100),
    fillOpacity = 0.7,
    # line
    dashArray   = "3",
    weight      = 2,
    color       = "white",
    opacity     = 1,
    # interaction
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labelsv,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(
    pal = palv, values = ~PERSONS_VACCINATED_1PLUS_DOSE_PER100, opacity = 0.7, title = HTML("Persons vaccinated 1+ doses per 100 population"),
    position = "bottomright")
setView(v_map,lng=-30, lat=-20, 1)
```

## A Time Series Analysis for Covid-19 Cases and Deaths

Covid-19 started at the beginning of 2020, and spread quickly accross the world. To analyze the overall spread pattern of the pademic, we analyze the global cases and deaths data as a time series. By plotting the time series by day, we observed a strong cyclic pattern, making the data very noisy. This is because of the delay of data update during weekends of the original dataset. To remove the cyclic pattern, it's better to use the weekly time series.

From the two weekly time series, we observed the following patterns.

* The first spike of reported deaths appeared in April 2020, with no spike in reported cases at the same time. This refer to the early spread in China. The low cases and high deaths also indicate the high fatality rate of the virus at the beginning.

* In January 2021, both the reported cases and reported deaths experience a relatively strong spike. During this time, the has pandemic spreaded worldwide, and contries in Europe and America are experiencing a high infection rate of Covid-19.

* The reported cases and deaths began decreasing slowly with vaccination being available since December 2020.

* The most outstanding spike of reported cases appeared in January 2022, with over 20 million weekly new cases around the world. This was over four times of that in January 2021. This was exactly the time when Omicron variant spread drastically accross the world. However, the report deaths did not increase that much at the mean time, from which we could conclude that Omicron spreads more easily than earlier variants, but causes less severe illness and death in general.

```{r}
# global new cases by day
p1 <- ggplot(world_daily,aes(Date_reported,world_new_cases))+
  geom_line(color='darkred')+
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Cases')+
  ggtitle('Global Daily New Cases')+
  theme_light()

# global new cases by week
p2 <- ggplot(world_weekly, aes(date_reported,weekly_new_cases))+
  geom_line(color='darkred')+
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Cases')+
  ggtitle('Global Weekly New Cases')+
  theme_light()

# global new deaths by day
p3 <- ggplot(world_daily, aes(Date_reported,world_new_deaths))+
  geom_line(color='steelblue')+
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('Deaths')+
  ggtitle('Global Daily New Deaths')+
  theme_light()

# global new deaths by week
p4 <- ggplot(world_weekly, aes(date_reported,weekly_new_deaths))+
  geom_line(color='steelblue')+
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('Deaths')+
  ggtitle('Global Weekly New Deaths')+
  theme_light()

p <- ggarrange(p1, p3, p2, p4,ncol = 2,nrow=2)
p
```


<font color='red'>  

Time series for US Cases & Deaths

Explanations here.

</font>


```{r echo=TRUE, message=FALSE, warning=FALSE}
p1 <- ggplot(us_daily, aes(x = Date_reported, y = New_cases)) +
  geom_line(color = "darkred") +
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Cases')+
  ggtitle('US Daily New Cases')+
  theme_light() +
  theme(legend.position="bottom",plot.title = element_text(hjust = 0.5))

  
p2 <- ggplot(us_daily, aes(x = Date_reported, y = New_deaths)) +
  geom_line(color = "steelblue") +
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Deaths')+
  ggtitle('US Daily New Deaths')+
  theme_light() +
  theme(legend.position="bottom",plot.title = element_text(hjust = 0.5))


p3 <- ggplot(us_weekly,aes(x = Date_reported, y = weekly_new_cases)) +
  geom_line(color = "darkred") +
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Cases')+
  ggtitle('US Weekly New Cases')+
  theme_light() +
  theme(legend.position="bottom",plot.title = element_text(hjust = 0.5))


p4 <- ggplot(us_weekly, aes(x = Date_reported, y = weekly_new_deaths)) +
  geom_line(color = "steelblue") +
  scale_x_date(breaks = function(x) seq.Date(from = min(x), to = max(x), by = "4 months"),
               labels = function(x) paste(lubridate::year(x),'-',lubridate::month(x), sep=''))+
  xlab('Date Reported')+
  ylab('New Deaths')+
  ggtitle('US Weekly New Deaths')+
  theme_light() +
  theme(legend.position="bottom",plot.title = element_text(hjust = 0.5))

p <- ggarrange(p1, p2, p3, p4,ncol = 2,nrow=2)
p

```

## Understanding the Vaccination in the US

In this section, we focus on the US Covid-19 vaccination data. By analyzing and visualizing the data, we try to answer the following questions - what is the overall level of US vaccination comparing to global data? How many Americans have been fully vaccinated while how many of them haven't taken any vaccination? Which state has the highest vaccination rate?

We first draw a boxplot of the fully vaccinated rate of different countries across the world. The median of global fully vaccinated rate is 61.165, while the US fully vaccinated rate is 67.989, which is between the median and third quantile of the global data.

```{r}
ggplot(world_vac, aes(x=fully_vaccinated_rate)) +
  geom_boxplot()+
  geom_segment(x=67.789,xend=67.789,y=-0.37,yend=0.37,color='red',linetype='dotted')+
  geom_text(aes(x=73, label="\nUnited States: 67.989", y=0.3), color="red",size=3)+
  ggtitle('US Fully Vaccinated Rate Comparing to Global Data') +
  xlab('Fully Vaccinated Rate')+
  ylab('')+
  theme(legend.title=element_blank())+
  coord_flip()+
  theme_light()
```

To understand the different vaccination state of the US population, we produced a stacked bar chart. we observed that 68% of the US population has been fully vaccinated, among which 33.7% of the population has taken the booster. However, there are still 20.3% of the population who have not taken any vaccine.

```{r}
us_vac$country <- rep('',4)
ggplot(us_vac, aes(country,value,fill=fct_rev(name)))+
  geom_col()+
  geom_text(aes(x=1, label="33.7%", y=26), color="black",size=4)+
  geom_text(aes(x=1, label="34.3%", y=60), color="black",size=4)+
  geom_text(aes(x=1, label="11.7%", y=74), color="black",size=4)+
  geom_text(aes(x=1, label="20.3%", y=94), color="black",size=4)+
  xlab('')+
  ylab('Proportion')+
  ggtitle('Vaccination State of US Population')+
  scale_fill_discrete_sequential(palette = "Terrain2")+
  theme(plot.margin=unit(rep(2,8),'cm'))+
  theme_light()+
  theme(legend.title=element_blank())+
  coord_flip()
```

A deeper insight of the vaccination in the US would be dividing the total population by age groups. From the grouped bar chart below, both the vaccinated and fully vaccinated rate grow with the age groups. People aged over 65 have the highest fully vaccinated rate of 88.6% and vaccinated rate of 95%. This is reasonable because the old people are more prone to be infected by the virus, so they are more likely to take the complete series of vaccines. Children less than 12 years are too young so their parents may worry about the side effects of the vaccine, so they have a lowest fully vaccinated rate of 16.7% and a vaccinated rate of 25.6%.

```{r}
ggplot(vac_age_long, aes(fct_relevel(age,'5 - 11 Years'),value,fill=name)) +
  geom_col(position=position_dodge(),alpha=0.6) +
  geom_text(aes(label=value), position=position_dodge(width=0.9), vjust=-0.3,size=3)+
  xlab('Ages')+
  ylab('Vaccinated Rate') +
  ggtitle('Vaccinatied and Reported Case Rate by Age')+
  scale_fill_discrete(name='')+ 
  theme_light()
```

<font color='red'>
Brief explanation here
</font>

```{r echo=TRUE, message=FALSE, warning=FALSE}
us_vac_density <- vaccinated_us %>%
  filter(state == 'United States') %>%
  mutate(color = 'red')

ggplot(vaccinated_us, aes(fully_vaccinated_rate)) +
  geom_density(fill = "pink",alpha = .3) +
  geom_vline(aes(xintercept = fully_vaccinated_rate),
             data = us_vac_density, color = 'red',linetype = 'dotted')+
  geom_text(aes(x = 75, label="\nUS Average: 68.9%", y=0.012), color="red")+
  ggtitle("US States Fully Vaccinated Rate") +
  theme(plot.title = element_text(hjust = .5))+
  xlab('Fully Vaccinated Rate')
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(vaccinated_us, aes(x = fully_vaccinated_rate)) +
  geom_boxplot()+
  geom_segment(x=68.9,xend=68.9,y=-0.37,yend=0.37,color='red',linetype='dotted')+
  geom_text(aes(x=71, label="\nUS Average: 68.9%", y=0.3), color="red",size=3)+
  ggtitle('US Average Fully Vaccinated Rate Comparing to States') +
  theme(legend.position="bottom", plot.title = element_text(hjust = .5)) +
  xlab('Fully Vaccinated Rate')+
  ylab('')+
  theme(legend.title=element_blank())+
  coord_flip()+
  theme_light()
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
theme_dotplot <- 
  theme_bw(16) +
  theme(axis.text.y = element_text(size = rel(.8)), axis.ticks.y = element_blank(),
        axis.title.x = element_text(), axis.text = element_text(face = "bold"),
        plot.background = element_rect(fill = "lightcyan2"),
        panel.background = element_rect(fill = "moccasin"),
        panel.grid.major.x = element_line(size = 0.5),
        panel.grid.major.y = element_line(size = 0.5, color = "lightblue"),
        panel.grid.minor.x = element_blank(),
        strip.text = element_text(size = rel(.7)), legend.position = "top")

vaccinated_us <- vaccinated_us |> 
  mutate(Panel = cut(fully_vaccinated_rate, 5, breaks = fivenum(fully_vaccinated_rate),
                     labels = c("<59.5%", "59.5% - 66.5%", "66.5% - 75.0%", ">76.5%")))  |> 
  mutate(Panel = fct_rev(Panel))

ggplot(vaccinated_us, aes(x = fully_vaccinated_rate, y = reorder(state, fully_vaccinated_rate))) +
  geom_point() +
  facet_wrap(~Panel, scales = "free") +
  ggtitle("% of People Fully Vaccinated by States") +
  theme(plot.title = element_text(hjust = .5))+
  xlab("% of People Fully Vaccinated") +
  ylab("") +
theme_dotplot
```

## How Does Vaccination Influence Covid-19 Cases?

<font color='red'>
Explanation here.

Feel free to change this section :)
</font>

```{r echo=TRUE, message=FALSE, warning=FALSE}
weekly_merge %>% filter(country_region == "US") %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = doses_n), color = "darkred") +
  labs(y = 'Doses', x = 'Months') +
  geom_line(aes(y = confirmed_n/0.025), color="steelblue", linetype="twodash") +
  scale_x_date(date_breaks = "3 month",date_labels = "%b%y") +
  scale_y_continuous(sec.axis = sec_axis(~.*0.025, name = "Cases")) +
  theme_light() +
  theme(legend.position="bottom",plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(name="Doses",
  values =  c("steelblue","darkred"),breaks = c("Doses", "Cases")) +
  ggtitle("Covid-19 Cases vs Doses in United States")
```

